Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-14 rigid_bodies.tex

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.

It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified one.
However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
In this paper, we devise a novel procedure that partially overcomes this limitation.
It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one. It should be emphasized that this approach is not limited to deterministic thermostats. It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of \citeauthor{Bussi_2007} \cite{Bussi_2007} and \citeauthor{Leimkuhler_2009} \cite{Leimkuhler_2009}, for instance.

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors. Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.

@@

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of \citeauthor{Tuckerman_2001} \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
We have done it by adapting the integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996} in light of a criterion explained by \citeauthor{Ezra_2006} \cite{Ezra_2006}.
It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.

@@

Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
\citeauthor{Omelyan_2008}'s processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a simpler path, though.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily by following \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(\timestep^n).
\end{equation}

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation*}
\label{eq:equipartition}
L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation*}
where subscript $\timestep$ points out the dependency of this average on the employed time step size.
The expression above extends the main result of \citeauthor{Eastwood_2010} \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that such temperature concerns the system that has actually been simulated, rather than the one initially specified.

@@

In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} with our described adaptations.

@@

Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
This corresponds to the angular coefficient of such quantity expressed as a function of time and is determined by least-square regression.
Recall that the symmetric scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
In the case of our proposed method, which we refer to as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations, with $18~\text{ns}$ of production time, carried out with different time-step sizes.
As one can observe, long-term stability is slightly improved by our method if $\timestep \le 5~\text{fs}$, indicating that there is little or no impact of the lack of time-reversal symmetry on $R$.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
	\label{fig:energy_drift}
\end{figure}

Fig.~\ref{fig:energy_drift} also contains drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in an NVE simulation \cite{Silveira_2017}.
Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme.
This indicates that the major causes for the observed drift took place during evaluations of the NVE propagator.
As the interaction potential we use is very smooth at cutoff, a possible cause is the amplification of round-off errors which can occur in events of close interatomic approaches \cite{Engle_2005}.
For time steps larger than $6~\text{fs}$, the NVE integration becomes fully unstable.
In contrast, each NVT integrator manages to remain numerically stable at least up to $h = 10~\text{fs}$, but at the cost of displaying an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
Recall that this part of $\refined H$ represents the total energy of the physical system.
Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, which shows that the thermostat stabilizes the numerical integration of the physical variables.
This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010} and also by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.

To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} is employed.
Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
Considering a different factorization scheme, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
%Nevertheless, this finding did not result in a revision of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in the literature.
This is confirmed for the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
In this figure, the negative energy values always correspond to the thermostat coupled to the translational degrees of freedom.
Therefore, energy flows from this thermostat to the one coupled to the rotational degrees.
Interestingly, Fig.~\ref{fig:num_stab} shows that the rate at which the thermostats approach the steady state decreases with $\timestep$. As far as we know, this behavior has not been reported before.

\begin{figure}
	\includegraphics{Figures/numerical_stability.eps}
	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}. In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom.}
	\label{fig:num_stab}
\end{figure}

@@

We have included the results of applying our method, but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.

Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, because that is not the only problematic issue.
Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}. Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
Thus, the results for the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.

@@

In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water. 
We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
From Fig.~\ref{fig:energy_partition} (d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to 5 fs.
For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.

>>>>

Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
This brings into question the NVT-MD integrator developed by Kamberaj \textit{et al}. \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As Eastwood \textit{et al}. \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.

It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified one.
However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
In this paper, we devise a novel procedure that partially overcomes this limitation.
It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one. It should be emphasized that this approach is not limited to deterministic thermostats. It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}. \cite{Bussi_2007} and Leimkuhler \textit{et al}. \cite{Leimkuhler_2009}, for instance.

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors. Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by Kamberaj \textit{et al}. \cite{Kamberaj_2005} and a reimplementation of the single-chain method by Martyna \textit{et al}. \cite{Martyna_1996} for rigid bodies.

@@

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}. \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
We have done it by adapting the integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996} in light of a criterion explained by Ezra \cite{Ezra_2006}.
It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.

@@

Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
Omelyan's processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
Here we choose to take a simpler path, though.
Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily by following Eastwood \textit{et al}. \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(\timestep^n).
\end{equation}

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation*}
\label{eq:equipartition}
L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation*}
where subscript $\timestep$ points out the dependency of this average on the employed time step size.
The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that such temperature concerns the system that has actually been simulated, rather than the one initially specified.

@@

In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} and the method of Martyna \textit{et al}. \cite{Martyna_1996} with our described adaptations.

@@

Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
This corresponds to the angular coefficient of such quantity expressed as a function of time and is determined by least-square regression.
Recall that the symmetric scheme of Martyna \textit{et al}. \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
In the case of our proposed method, which we refer to as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations, with $18~\text{ns}$ of production time, carried out with different time-step sizes.
As one can observe, long-term stability is slightly improved by our method if $\timestep \le 5~\text{fs}$, indicating that there is little or no impact of the lack of time-reversal symmetry on $R$.

\begin{figure}
	\includegraphics{Figures/energy_drift.eps}
	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to Martyna \textit{et al}. \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
	\label{fig:energy_drift}
\end{figure}

Fig.~\ref{fig:energy_drift} also contains drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in an NVE simulation \cite{Silveira_2017}.
Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme.
This indicates that the major causes for the observed drift took place during evaluations of the NVE propagator.
As the interaction potential we use is very smooth at cutoff, a possible cause is the amplification of round-off errors which can occur in events of close interatomic approaches \cite{Engle_2005}.
For time steps larger than $6~\text{fs}$, the NVE integration becomes fully unstable.
In contrast, each NVT integrator manages to remain numerically stable at least up to $h = 10~\text{fs}$, but at the cost of displaying an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
Recall that this part of $\refined H$ represents the total energy of the physical system.
Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, which shows that the thermostat stabilizes the numerical integration of the physical variables.
This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkhuler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.

To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} is employed.
Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
As already mentioned, the scheme due to Kamberaj \textit{et al}. \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
Considering a different factorization scheme, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
%Nevertheless, this finding did not result in a revision of the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} in the literature.
This is confirmed for the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
In this figure, the negative energy values always correspond to the thermostat coupled to the translational degrees of freedom.
Therefore, energy flows from this thermostat to the one coupled to the rotational degrees.
Interestingly, Fig.~\ref{fig:num_stab} shows that the rate at which the thermostats approach the steady state decreases with $\timestep$. As far as we know, this behavior has not been reported before.

\begin{figure}
	\includegraphics{Figures/numerical_stability.eps}
	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by Kamberaj \textit{et al}. \cite{Kamberaj_2005}. In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom.}
	\label{fig:num_stab}
\end{figure}

@@

We have included the results of applying our method, but using the approximate solution of rotation due to Miller \textit{et al}. \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.

Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, because that is not the only problematic issue.
Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al}. \cite{Martyna_1996}. Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
Thus, the results for the method of Martyna \textit{et al}. \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.

@@

In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water. 
We see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of Martyna \textit{et al}. \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
From Fig.~\ref{fig:energy_partition} (d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to 5 fs.
For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}. \cite{Martyna_1996} and  Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
